The Impact of Alternation

### 2017-05-15 draft.tex
ok

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{blast} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language of the automaton, if one exists, but
termination is not guaranteed when the language is empty.

@@

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{blast}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the decision procedure.

The emptiness check for alternating data automata adapts similar
semi-algorithms for nondeterministic infinite-state programs to the
alternating model of computation. In particular, we considered the
state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
outperform lazy predicate abstraction \cite{blast} in the
nondeterministic case, and generalized it to cope with
alternation. More recent approaches for interpolant-based abstraction
refinement target Horn systems \cite{McMillan14,Hoder12}, used to
encode recursive and concurrent programs
\cite{Grebenshchikov12}. However, the emptiness of alternating word
automata cannot be directly encoded using Horn clauses, because all
the branches of the computation synchronize on the same input, which
cannot be encoded by a finite number of local (equality)
constraints. We believe that the lazy annotation techniques for Horn
clauses are suited for branching computations, which we intend to
consider in a future tree automata setting.

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
only atoms $x \geq k$ and $x = \overline{x} + k$, for $k \in \zed$,
with the classical interpretation of the function symbols on
integers. Since reachability of a control state is undecidable for
AVASS \cite{LINCOLN92}, the emptiness problem is undecidable for ADA.

@@

We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
to the total lexicographical order $\prec^*$ on $\Sigma^*$. A node $n
\in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an ancestor $m$
such that $(m,p) \in \lhd$, for some $p \in N$. A node $n$ is
\emph{closed} if it is covered, or $\Lambda(n) \not\models \Lambda(m)$
for all $m \in N$ such that $\lambda(m) \prec^* \lambda(n)$. Observe
that we use the coverage here with a different meaning than in
Algorithm \ref{alg:predabs}.

@@

\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
  \setminus L(\A_1)$. \qed}

@@

\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
  and $i \in [0,n]$ be the first index for which $I_i \not\in
  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
  an interpolant, because: \begin{compactitem}
  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
    %
  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
  \end{compactitem}
  We proceed with these replacements until there are no more formulae
  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}

@@

  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
  interpolant for $\pi(u)$, the following entailments
  hold: \begin{compactitem}
  \item $\theta_0 \models I_0[Q_0/Q]$,
    %
  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
    %
  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
  \end{compactitem}
  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
  induction on $n \geq 0$. This is sufficient to conclude because
  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
  \models I_0$. For the induction step $n>0$, we compute:
  \[\begin{array}{l}
  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q]
  \end{array}\]
\qed}

@@

\proof{We prove the following invariant: each time Algorithm
  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
  in $\worklist$ contains all the frontier nodes in the ART
  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
  some node in $N$, namely that:
  \begin{equation}\label{inv:predabs}
  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
    \not\in E ~\wedge~ (n,m) \not\in \lhd}
  \end{equation}
  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
  = \lhd = \emptyset$. If the invariant holds previously, at line
  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
  executed, because, when the subtree rooted at the pivot $p$ is
  removed, $p$ becomes a member of the set of uncovered frontier
  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
  considers that the invariant holds at line \ref{ln:while} and the
  control follows the else branch at line \ref{ln:else}. In this case,
  the newly created frontier node $s$ is added to $W$ only if it is
  not covered by an existing node in $N$ (line \ref{ln:covered}).

@@

  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
  invariant, suppose, by contradiction, that there exists $u \in
  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
  is satisfiable as well. However, this cannot be the case, because
  $p$ has been processed at line \ref{ln:feasible} and Algorithm
  \ref{alg:predabs} would have returned a counterexample,
  contradicting the assumption that it returns $\true$. This concludes
  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}.

  We have then proved the second point of the statement. For the first
  point, assume that $L(\A) \neq 0$ and let $w = (a_1,\nu_1) \ldots
  (a_k,\nu_k) \in L(\A)$ be a word. By the above, Algorithm
  \ref{alg:predabs} cannot return $\true$. Suppose, by contradiction
  that it does not terminate. Since the sequences from $\Sigma^*$ are
  explored in breadth-first order, every sequence of length $k$ is
  eventually explored, which leads to the discovery of $w$ at line
  \ref{ln:feasible}. Then Algorithm \ref{alg:predabs} terminates
  returning $w \in L(\A)$. \qed}

@@

\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
  the entailment holds trivially. Otherwise, let $\Omega$ be the set
  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
  be a spurious counterexample. Let
  $\langle\top,I^\omega_0,\ldots,I^\omega_k\equiv\bot\rangle$ be an
  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
  \ref{prop:positive-interpol}, it is possible to build such an
  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
  Proposition \ref{prop:asterisk}, we obtain
  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}

@@

\proof{ We prove first that, each time Algorithm \ref{alg:impact}
  reaches the line \ref{ln:impact-while}, we have:
  \begin{equation}\label{inv:impact}
  W = \set{n \not\in N \mid n \text{ uncovered, } \exists a \in
    \Sigma~ \forall s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus
  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
  holds at when reaching line \ref{ln:impact-while} and some node $n$
  was removed from $W$ and inserted into $N$. We distinguish two
  cases, either: \begin{itemize}
  \item $n$ is covered, in which case $W$ becomes $W \setminus
    \set{n}$, or
  %
  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
      a \in \Sigma}$ is the set of fresh successors of $n$. But then
    no node $s \in S$ is covered and has successors in $E$.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time line
  \ref{ln:impact-while} is reached, thus it is invariant.

>>>>

The boolean operations of union, intersection and complement can be
implemented in linear time in this model, thus matching the complexity
of performing these operations in the finite-alphabet case. The price
to be paid is that emptiness becomes undecidable, for which reason we
provide two efficient semi-algorithms for emptiness, based on lazy
predicate abstraction \cite{HJMS02} and the \impact~method
\cite{mcmillan06}.  These algorithms are proven to terminate and
return a word from the language of the automaton, if one exists, but
termination is not guaranteed when the language is empty.

@@

The model presented in this paper stems from the language inclusion
problem considered in \cite{Tacas16}. There we provide a
semi-algorithm for inclusion of data languages, based on an
exponential determinization procedure and an abstraction refinement
loop using lazy predicate abstraction \cite{HJMS02}. In this work we
consider the full model of alternation and rely entirely on the
ability of SMT solvers to produce interpolants in the combined theory
of booleans and data. Since determinisation is not needed and
complementation is possible in linear time, the bulk of the work is
carried out by the decision procedure.

The emptiness check for alternating data automata adapts similar
semi-algorithms for nondeterministic infinite-state programs to the
alternating model of computation. In particular, we considered the
state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
outperform lazy predicate abstraction \cite{HJMS02} in the
nondeterministic case, and generalized it to cope with
alternation. More recent approaches for interpolant-based abstraction
refinement target Horn systems \cite{McMillan14,Hoder12}, used to
encode recursive and concurrent programs
\cite{Grebenshchikov12}. However, the emptiness of alternating word
automata cannot be directly encoded using Horn clauses, because all
the branches of the computation synchronize on the same input, which
cannot be encoded by a finite number of local (equality)
constraints. We believe that the lazy annotation techniques for Horn
clauses are suited for branching computations, which we intend to
consider in a future tree automata setting.

@@

Unlike the boolean closure properties, showed to be effectively
decidable (Lemma \ref{lemma:closure}), the emptiness problem is
undecidable, even in very simple cases. For instance, if $\Data^\I$ is
the set of positive integers, an ADA can simulate an Alternating
Vector Addition System with States (AVASS) using only atoms $x \geq k$
and $x = \overline{x} + k$, for $k \in \zed$, with the classical
interpretation of the function symbols on integers. Since reachability
of a control state is undecidable for AVASS \cite{LINCOLN92}, the
emptiness problem is undecidable for ADA.

@@

We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
to the total lexicographical order $\prec^*$ on $\Sigma^*$. A node $n
\in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an ancestor $m$
such that $(m,p) \in \lhd$, for some $p \in N$. A node $n$ is
\emph{closed} if it is covered, or $\Lambda(n) \not\models \Lambda(m)$
for all $m \in N$ such that $\lambda(m) \prec^* \lambda(n)$. Observe
that we use the coverage relation $\lhd$ here with a different meaning
than in Algorithm \ref{alg:predabs}.

@@

\vspace*{\baselineskip}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)} \enspace.
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
  L(\overline{\A_1})$. \qed}

@@

\vspace*{\baselineskip} \proof{ Let
  $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant, and $i \in
  [0,n]$ be the first index for which $I_i \not\in
  \Form^+(Q,\vec{x})$. If $i=0$, we replace $I_0$ with $\iota \in
  \Form^+(Q,\vec{x})$ and verify that the result is still an
  interpolant. If $i>0$, by Proposition \ref{prop:asterisk}, we have
  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
  an interpolant, because: \begin{compactitem}
  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
    %
  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
  \end{compactitem}
  We proceed with these replacements until there are no more formulae
  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}

@@

  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
  interpolant for $\Theta(u)$, the following entailments
  hold: \begin{compactitem}
  \item $\theta_0 \models I_0[Q_0/Q]$,
    %
  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
    %
  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
  \end{compactitem}
  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
  induction on $n \geq 0$. This is sufficient to conclude because
  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
  \models I_0$. For the induction step $n>0$, we compute:
  \[\begin{array}{l}
  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
  \end{array}\]}

@@

\proof{We prove the following invariant: each time Algorithm
  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
  in $\worklist$ contains all the frontier nodes in the ART
  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
  some node in $N$, namely that:
  \begin{equation}\label{inv:predabs}
  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
    \not\in E ~\wedge~ (n,m) \not\in \lhd}
  \end{equation}
  Initially, this is the case because $W = \set{\rootNode}$ and $E =
  \lhd = \emptyset$. If the invariant holds previously, at line
  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
  executed, because, when the subtree rooted at the pivot $p$ is
  removed, $p$ becomes a member of the set of uncovered frontier
  nodes, and is added to $W$ at line \ref{ln:refine}. Otherwise, the
  invariant holds at line \ref{ln:while} and the control follows the
  else branch at line \ref{ln:else}. In this case, the newly created
  frontier node $s$ is added to $W$ only if it is not covered by an
  existing node in $N$ (line \ref{ln:covered}).

@@

  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
  invariant, suppose, by contradiction, that there exists $u \in
  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
  is satisfiable as well. However, this cannot be the case, because
  $p$ has been processed at line \ref{ln:feasible} and Algorithm
  \ref{alg:predabs} would have returned a counterexample,
  contradicting the assumption that it returns $\true$. This concludes
  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}. We
  have then proved the second point of the statement.

  For the first point, assume that $L(\A) \neq 0$ and let $w =
  (a_1,\nu_1) \ldots (a_k,\nu_k) \in L(\A)$ be a word. By the above,
  Algorithm \ref{alg:predabs} cannot return $\true$. Suppose, by
  contradiction that it does not terminate. Since the sequences from
  $\Sigma^*$ are explored in breadth-first order, every sequence of
  length $k$ is eventually explored, which leads to the discovery of
  $w$ at line \ref{ln:feasible}. Then Algorithm \ref{alg:predabs}
  terminates returning $w \in L(\A)$. \qed}

@@

\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
  the entailment holds trivially. Otherwise, let $\Omega$ be the set
  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
  be a spurious counterexample. Let
  $\langle\top,I^\omega_0,\ldots,I^\omega_k,\bot\rangle$ be an
  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
  \ref{prop:positive-interpol}, it is possible to build such an
  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
  Proposition \ref{prop:asterisk}, we obtain
  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}

@@

\proof{ We prove first that, each time Algorithm \ref{alg:impact}
  reaches the line \ref{ln:impact-while}, we have:
  \begin{equation}\label{inv:impact}
  W = \set{n \mid n \text{ uncovered, } \exists a \in \Sigma~ \forall
    s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $E=\lhd=\emptyset$, thus
  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
  holds at when reaching line \ref{ln:impact-while} and some node $n$
  was removed from $W$ and inserted into $N$. We distinguish two
  cases, either: \begin{itemize}
  \item $n$ is covered, in which case $W$ becomes $W \setminus
    \set{n}$ and (\ref{inv:impact}) holds, or
  %
  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
      a \in \Sigma}$ is the set of fresh successors of $n$. But then
    no node $s \in S$ is covered and has successors in $E$, thus
    (\ref{inv:impact}) holds.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time line
  \ref{ln:impact-while} is reached, thus it is invariant.
